Atmospheric escape and star-planet-interaction in the HAT-P-67 b system

### 2023-02-07 paper/paper1/ms.tex
Retool the introduction

Atmospheric escape appears to be a compulsory phase for exoplanets of a certain size and insolation, statistically imprinted in the dearth of 1.5-2.0~$R_\oplus$ planets in the California-Kepler survey \citep{2017AJ....154..109F}.  The two physical mechanisms for this fleeting interlude between mini-Neptunes and super-Earths include photoevaporation \citep{2013ApJ...775..105O,2017ApJ...847...29O} and core-powered mass loss \citep{2019MNRAS.487...24G}.  Whatever the cause, some large fraction of planets must undergo atmospheric escape and the signal should be widely discernable, at least in observationally favorable configurations.

The most amenable observing configuration involves transit spectroscopy in one of three key atomic transitions: Ly~$\alpha$, \ion{He}{1} 10833 \AA, or H~$\alpha$...

Enough uncertainty exists in the tradeoffs among physical mechanisms, system ages, and timescales to make it unclear whether any given exoplanet will be actively undergoing atmospheric escape.  The most favorable conditions may include certain commonalities: close-in, low surface gravity exoplanets with a strategically chosen radius range.  The 1.5-2.0 Earth radius range could be expected to harbor atmospheres that have not yet shed their gaseous envelopes, still in the act of descending to their final state.  Alternatively, inflated Hot Jupiters could be so dramatically ablated as to overcome the somewhat stronger potential well.  Significant gains in understanding can therefore be expected from observational studies sampling the extremes of the exoplanet distribution.

8 systems have Helium 10830 exosphere measurements to date: \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.

Here we present a study of the exosphere of HAT-P-67b.  HAT-P-67 is an F type sub-giant with the Saturn-sized planet HAT-P-67b.  HAT-P-67b has a low surface gravity, which makes it amenable to rapid atmospheric mass loss.  It orbits an evolved star, which means its instellation may have increased as the star's size has grown.  The star appears close to the star, making it conceivable that star-planet interactions could be at play.  These and other reasons make HAT-P-67b a unique laboratory for measuring the properties of exospheres and their interactions with their host stars.  The system appears to consist of a binary with an M dwarf companion HAT-P-67B separated by $9\farcs0$.

>>>>

Atmospheric escape appears to be a compulsory phase for exoplanets of a certain size and insolation, statistically imprinted in the dearth of 1.5-2.0~$R_\oplus$ planets in the California-Kepler survey \citep{2017AJ....154..109F}.  The two physical mechanisms for this fleeting interlude between mini-Neptunes and super-Earths include photoevaporation \citep{2013ApJ...775..105O,2017ApJ...847...29O} and core-powered mass loss \citep{2019MNRAS.487...24G}.  Whatever the cause, some large fraction of planets must undergo atmospheric escape and the signal should be widely discernable, at least in observationally favorable configurations such as transiting planets.

Enough uncertainty exists in the tradeoffs among physical mechanisms, system ages, and timescales to make it unclear whether any given exoplanet will be actively undergoing atmospheric escape.  The most favorable traits may include close-in, low surface gravity exoplanets within a strategically chosen radius range.  The 1.5-2.0 Earth radius range could be expected to harbor atmospheres that have not yet shed their gaseous envelopes, still in the act of descending to their final state.  Alternatively, inflated Hot Jupiters could be so dramatically ablated as to overcome the somewhat stronger potential well.  Significant gains in understanding can therefore be expected from observational studies sampling the extremes of the exoplanet distribution.

Candidates for detection should be observable in one of three atomic lines.  Ly~$\alpha$, \ion{He}{1} 10833 \AA, and H~$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2023MNRAS.518.4357O}.  Ly $\alpha$ can exhibit strong atmospheric escape signals, but instellar medium \ion{H}{1} absorption censors its low-velocity linecore.  \ion{He}{1} 10833 \AA offers significant advantages over Ly~$\alpha$, and has resulted in 8 systems with detections to date: \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  These detections exhibit a range of signal strengths, as high as 10$\%$ transit depths.  Measurements at high spectral resolution have been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.

Here we present a study of the exosphere of \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant\footnote{The system appears to consist of a binary with an M dwarf companion HAT-P-67B separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}.  The IAU naming convention would demand HAT-P-67Ab.  We simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.} at a distance of 0.06 AU \citep{2017AJ....153..211Z}.  The strong insolation combined with \object{HAT-P-67 b}'s low surface gravity makes it an excellent candidate for strong atmospheric mass loss.  It orbits an evolved star, opening the prospect for the planet's insolation to have increased over time as the star grew.  These and other reasons make \object{HAT-P-67 b} a predestined laboratory for measuring the properties of exospheres at the extreme.
